---
title: "p8105_hw2_mk4022"
author: "Mary Knoop"
date: "10/2/2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Load libraries for analyses 
```{r}
library(tidyverse)
library(readxl)
library(knitr)
```

Problem 1
Part A:
* specify the sheet in the Excel file and to omit non-data entries (rows with notes / figures; columns containing notes) using arguments in read_excel
* use reasonable variable names
* omit rows that do not include dumpster-specific data
* round the number of sports balls to the nearest integer and converts the result to an integer variable (using as.integer)

```{r}
trash_wheel = 
  read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = 1, skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na(dumpster)%>% 
  select(-x15, -x16, -x17) %>% 
  mutate(sports_balls = as.integer(sports_balls))

#view datasets for correctness 
trash_wheel
```


Part B:
Read and clean precipitation data for 2017 and 2018. For each, omit rows without precipitation data and add a variable year.

```{r}
precipitation_17 = 
  read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = 6, skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na() %>% 
  mutate(year = 2017) %>% 
  select(year, month, total)

precipitation_18 = 
  read_excel("./data/Trash-Wheel-Collection-Totals-8-6-19.xlsx", sheet = 5, skip = 1) %>% 
  janitor::clean_names() %>% 
  drop_na() %>% 
  mutate(year = 2018) %>% 
  select(year, month, total)

#view datasets for correctness 
precipitation_17
precipitation_18
```

Part C:
Next, combine precipitation datasets and convert month to a character variable (the variable month.name is built into R and should be useful).
```{r}
precipitation_all = 
  full_join(precipitation_17, precipitation_18) %>% 
  mutate(month = month.name[month])

#view datasets for correctness 
precipitation_all
```

Part D: 
Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in both resulting datasets, and give examples of key variables. For available data, what was the total precipitation in 2018? What was the median number of sports balls in a dumpster in 2017?

The trash_wheel dataset has `r nrow(trash_wheel)` observations, and the precipitation_all data set has `r nrow(precipitation_all)` observations. Key variables in the dataset trash_wheel include `glass_bottles` and `grocery_bags`. Key variables in the dataset precipitation_all includes `year`, `month`, and `total`. The total precipitation in 2018 was `r sum(pull(precipitation_18, total))` inches. The median number of `sports_balls` in the dumpster in 2017 was `r median(pull(filter(trash_wheel, year==2017), sports_balls))`.

Problem 2
Part A: 
First, clean the data in pols-month.csv. Use separate() to break up the variable mon into integer variables year, month, and day; replace month number with month name; create a president variable taking values gop and dem, and remove prez_dem and prez_gop; and remove the day variable.

```{r}
pols =
  read_csv("./data/pols-month.csv") %>%
  janitor::clean_names() %>% 
  separate(mon, c("year", "month", "day")) %>%
    mutate(
    year = as.numeric(year),
    month = as.numeric(month),
    month = month.name[month]
    ) %>%
  mutate(president = ifelse(prez_gop == 1, "gop", "dem")) %>% 
  mutate(year = as.numeric(year)) %>% 
  select (-prez_gop, -prez_dem, -day)

#view datasets for correctness 
pols
```

Part B:
Second, clean the data in snp.csv using a similar process to the above. For consistency across datasets, arrange according to year and month, and organize so that year and month are the leading columns.

```{r}
snp =
  read_csv("./data/snp.csv") %>%
  janitor::clean_names() %>% 
  separate(date, c("month", "day", "year")) %>% 
  mutate(
    year = as.numeric(year),
    month = as.numeric(month),
    month = month.name[month]
    ) %>%
  select (year, month, -day, close) %>% 
  arrange (year, month)

#view dataset for correctness
snp
```

Part C: 
Third, tidy the unemployment data so that it can be merged with the previous datasets. This process will involve switching from “wide” to “long” format; ensuring that key variables have the same name; and ensuring that key variables take the same values.

```{r}
unemploy=
  read_csv("./data/unemployment.csv") %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    jan:dec, 
    names_to = "month", 
    values_to = "rate")

merge =
  left_join(pols, snp, by = c("year","month"))

final = 
  left_join(merge,unemploy, by =c("year","month"))

#view for correctness
unemploy
```
Part D:
Join the datasets by merging snp into pols, and merging unemployment into the result.
```{r}
merge =
  left_join(pols, snp, by = c("year","month"))

final = 
  left_join(merge,unemploy, by =c("year","month"))

#view for correctness
final
```

Part E: 
Write a short paragraph about these datasets. Explain briefly what each dataset contained, and describe the resulting dataset (e.g. give the dimension, range of years, and names of key variables).

The pol dataset shows who was in power from 1947 to 2015. The dataset snp shows S&P 500 data from 1950 to 2015. The dataset unemploy shows unemployment data from 1948 to 2015. The database final combines all these datasets and includes key variables such as `sen_dem`, `sen_gop`,`prez`, and `snp`. The database final  has `r nrow(final)` observations between 1947 to 2015. 

Problem 3
Part A:
Load and tidy the data. Note that, although these data may seem fairly well formatted initially, the names of a categorical predictor and the case structure of string variables changed over time; you’ll need to address this in your data cleaning. Also, some rows seem duplicated, and these will need to be removed (hint: google something like “dplyr remove duplicate rows” to get started).

```{r}
baby_names=
  read_csv("./Data/Popular_Baby_Names.csv") %>%
  janitor::clean_names() %>%
  distinct() %>%
  mutate(
    ethnicity = replace(ethnicity, ethnicity == "ASIAN AND PACI", "ASIAN AND PACIFIC ISLANDER"),
    ethnicity = replace(ethnicity, ethnicity == "BLACK NON HISP", "BLACK NON HISPANIC"),
    ethnicity = replace(ethnicity, ethnicity == "WHITE NON HISP", "WHITE NON HISPANIC"), 
         ethnicity = str_to_lower(ethnicity), gender = str_to_lower(gender), childs_first_name = str_to_lower(childs_first_name)) 

baby_names
```

Part B:
Produce a well-structured, reader-friendly table showing the rank in popularity of the name “Olivia” as a female baby name over time; this should have rows for ethnicities and columns for year.
```{r}
baby_names %>%
  filter(childs_first_name == "olivia") %>%
  select(-count) %>%
  pivot_wider(names_from = year_of_birth, values_from = rank)
```


Part C: 
Produce a similar table showing the most popular name among male children over time.
```{r}
baby_names %>%
  filter(gender == "male", rank == 1) %>%
  select(-count) %>%
  pivot_wider(names_from = year_of_birth, values_from = rank)
```

Part D:
Finally, for male, white non-hispanic children born in 2016, produce a scatter plot showing the number of children with a name (y axis) against the rank in popularity of that name (x axis).
```{r}
baby_names_scatterplot=
  filter(baby_names, gender == "male", ethnicity == "white non hispanic", year_of_birth == 2016) 
ggplot(baby_names_scatterplot, aes(x= rank, y= count)) + geom_point()
```

